---
title: "Predicting Fire Burn Occurrences"
author: ' Lina Nguyen,Cole Bailey,Kevin Stewart'
date: "12/3/2021"
output:
  word_document: default
  html_document: default
  pdf_document: default
---


# Importing packages
```{r}
library(tidyverse);library(tidyr);library(zoo);library(zoo);library(xts);library(tseries);library(astsa);library(lubridate);library(ggplot2);library(dplyr);library(fpp2);library(dplyr);library(tsibble);library(tseries);library(forecast);library(sarima);library(fpp2);library(ggplot2)

```


#loading the data file and seperating string values
```{r}
df_fire <- read.csv("/Users/kevinstewart/Desktop/business_a/san_diego_fire_incident.csv", sep = ",")
```


#viewing data
```{r}
head(df_fire)
```
# correct colnames & and subsetting data based on year and most occurring problem 
```{r}
names(df_fire) <- df_fire %>% slice(1) %>%
unlist()
fire <- df_fire %>% slice(-1)

```

# Mutate the data to month response
```{r}
fire$problem_count <- as.numeric(fire$problem_count)
fire$month_response <- as.numeric(fire$month_response)
fire$year_response <- as.numeric(fire$year_response)
fire$problem <- as.factor(fire$problem)

########################################
# verify structure of data 
str(fire)
```


# Verify the length and dimensions of the data 
```{r}
################################
# Instantiate the variable
burns_fire <- fire %>%
  filter(problem == "Burns / Explosion (L3)")

```


# Viewing frequency of data
```{r}
#############################################
ggplot(data = fire) + geom_histogram(mapping = aes(x = month_response)) + scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) + ggtitle('Number of Fire And EMS Dispatchment per Month') + xlab('Month') + ylab('Count')
#############################################
ggplot(data = fire) + geom_bar(mapping = aes(x = year_response)) + ggtitle("Number of Fire And EMS Dispatchment per Year") + xlab("Year") + ylab("Count") + scale_x_continuous(breaks = scales::pretty_breaks(n = 16))
##############################################
ggplot(data = burns_fire) + geom_bar(mapping = aes(x = month_response)) + scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) + ggtitle('Number of Fire Burn And Incidences per Month') + xlab('Month') + ylab('Count')
################################################
ggplot(data = burns_fire) + geom_bar(mapping = aes(x = year_response)) + scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) + ggtitle('Number of Fire Burn And Explosions per Year') + xlab('Year') + ylab('Count')
```


# Looking at time series data 
```{r}
par(mfrow=c(3:2))
tsplot(fire[,4], ylab = "", xlab = "", type = "l", main = "Times Series Plot of EMS Incidences", col = 4)
tsplot(fire[,4], xlab = "year_response", ylab = "Problem Count", main = "Time Series Plot of Fire Burns and Explosions", type = "l", lwd=2, xlim = c(2006, 2021))
tsplot(fire[,4], xlab = "", ylab = "", main = "Time Series Plot of Problem Count", type = "l", lwd=2, xlim = c(1,612))
```

# view data by individual fire incidents
```{r}
tsplot(burns_fire$month_response, main = "Time Series Plot of Fire Burns And Explosions")

# logging the data graphically
tsplot(log(burns_fire$month_response), main = "Time Series Plot of Logarithm Fire Burns and Explosions")

# the data shows that there is a there is also seasonality in the data which will need to be removed.
########################################
#viewing first variables of the burns month response data 
head(burns_fire$month_response)
#####################################
# graphically displaying month response and problem count
tsplot(fire$month_response, main = "Time Series Plot of the Month")
tsplot(fire$problem_count, main = "Time Series Plot of the Month Response and Problem Count")
```


# Declaring data as time series 
```{r}
# creating a time series variable
Y = ts(burns_fire[,4],start = c(2006,1),frequency = 12)
###############################################
# create a time series object to check for seasonality 
ts_fire_b <- ts(burns_fire[,5],start = c(2006,1),frequency = 12)

# Checking seasonality with the seasonal graph 
ggseasonplot(Y) + ggtitle("Seasonal Plot: Change in Monthly Burn and Explosion Incidents")

###############################################
ggsubseriesplot(ts_fire_b) +
  ggtitle("Seasonal Plot: Monthly Seasonal Graph of Fire and Explosion Incidences")+ ylab("Number of Burn Incidents")

#There shows that there is a spike in the winter to spring months and drastic decrease at the start of summer. 

```


# Looking at acf and pacf plot to check autocorrelation
```{r}
head(acf2(burns_fire$month_response,main =  "ACF and PACF of Monthly Burns and Explosions"))
#this show a significant amount of correlation in 

```


#Differencing the data to obtain stationarity to remove trend and seasonality
```{r}
diff_burns <- diff(diff(burns_fire$month_response))
tsplot(diff_burns)
# acf and pacf plots
acf2(diff_burns, main = "ACF and PACF of Differenced Monthly Burns and Explosions")
```


# Trying different AR models to find the best model 
```{r}
# Looking at ARIMA (1,1,1) model
ar1 <- arima(ts_fire_b, order = c(1,1,1))
ar1
# AIC of 262.73
###################################
# ARIMA (2,0,0)
ar2 <- arima(ts_fire_b, order = c(2,0,0))
ar2
###################################
# In viewing the data it appears that the data is best with differncing 
ar3 <- arima(ts_fire_b, order = c(3,0,0))
ar3
###################################
ar4 <- arima(ts_fire_b, order =c(4,0,0))
ar4
####################################
ar5 <- arima(ts_fire_b, order = c(5,0,0))
ar5

print(ar1);ar2;ar3;ar4;ar5
####################################
# It is determined that an ARIMA model of (4,0,0) is the best model for forecasting 
```


# Find the best arima model using a fit arima 
```{r}
# Determining best model
fit_ar <- auto.arima(diff_burns, stepwise = FALSE, approximation = FALSE, trace = TRUE)
###################################
# Confirming that an ARIMA (4,0,0) is the best fit model
```


#check residuals for the best model 
```{r}
#check residuals using forecast and a lag
forecast::checkresiduals(fit_ar, lag=12)
# the best model based on forecasting and checking residuals
arima(burns_fire$month_response, order = c(4,0,0)) 

# print summary statistics
checkresiduals(fit_ar)
print(summary(fit_ar))

#Get the standard deviation
std <- sqrt(17.64)
std

```

# In viewing the acf most of all the autocorrelation is removed from the model.


# Dickey fuller Test to check for stationarity
```{r}
adf.test(diff_burns) 
#The stationarity is removed according to the dickey fuller test 
```
# We would reject the null, shows that we have stationarity.


# Using arima to forecast monthly trends of fire incidents in San Diego
```{r}
# Forecasting 12 months into the future
f_cast <- forecast(fit_ar, h=12)
####################################
autoplot(f_cast)
# forecasting to look at most recent data
autoplot(f_cast, include = 48)

head(print(summary(f_cast)))

```








